Associations between Chlorophyll \emph{a} and various Microcystin-LR Health Advisory Concentrations

### 2015-09-11 manuscript/manuscript.Rmd
figures closer

```{r chla_micro_scatter, echo=FALSE, fig.width=8, fig.cap="Scatterplot showing association betweeen chlorophyll \\it{a} and microcystin-LR. \\label{fig:chla_micro_scatter}"}
plot_scatter(nla_dat,CHLA,MCYST_TL_UGL,viridis(1))
```

@@

```{r chla_micro_cdf, echo=FALSE, fig.width=8, fig.cap="Conditional probability plots showing association between the probability of exceeding various microcystin-LR (MLR) health advisory Levels. \\label{fig:chla_micro_cdf}"}
plot_cdf()
```

>>>>

```{r chla_micro_scatter, echo=FALSE, fig.width=8, fig.cap="Scatterplot showing association betweeen chlorophyll \\it{a} and microcystin-LR. \\label{fig:chla_micro_scatter}"}
plot_scatter(nla_dat,xvar="log_chla",yvar = "log_micro",pt_col=viridis(1),
             x=expression(paste('Log10 Chl ', italic("a"),' (',mu,'g/L)')),
             y=expression(paste('Log10 microcystin-LR',' (',mu,'g/L)')))
```

@@

```{r chla_micro_cdf, echo=FALSE, fig.width=8, fig.cap="Conditional probability plots showing association between the probability of exceeding various microcystin-LR (MLR) health advisory Levels. \\label{fig:chla_micro_cdf}"}
plot_cdf(nla_dat[,c("log_chla","log_micro")],"")
```

### 2015-09-11 manuscript/manuscript.tex
figures closer

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use and managing water bodies to meet
those levels could have far reaching benefits. However, compared to
other water quality measures, measurements of microcystin are not common
and current field measurement techniques have limited precision and
accuracy. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 20.8, 71.2, 92.16, and 110.16, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

We evaluated associated chlorophyll \emph{a} concentrations for an
effect for each of the WHO and U.S. EPA levels (Table
\ref{tab:microcystin_levels}). Lakes with higher microcystin-LR
concentrations were rare. Only 1.16 \% of lakes sampled had a
concentration greater than 10. For this analysis we focus on the
microcystin concentrations that are better represented in the NLA data.
These were 0.3, 1, 1.6, and 2 \(\mu\)g/L.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/epa_child_cp_plot-1.pdf}
\caption{Plot of \label{fig:epa_child_cp_plot}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/who_drink_cp_plot-1.pdf}
\caption{Plot of \label{fig:who_drink_cp_plot}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/epa_adult_cp -1.pdf}
\caption{Plot of \label{fig:epa_adult_cp}}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/who_rec_low1_cp-1.pdf}
\caption{Plot of \label{fig:who_rec_low1_cp}}
\end{figure}

@@

\begin{longtable}[c]{@{}lll@{}}
\toprule
\begin{minipage}[b]{0.11\columnwidth}\raggedright\strut
Source
\strut\end{minipage} &
\begin{minipage}[b]{0.16\columnwidth}\raggedright\strut
Type
\strut\end{minipage} &
\begin{minipage}[b]{0.19\columnwidth}\raggedright\strut
Concentration
\strut\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
1 ug/l
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
U.S. EPA
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
0.3 ug/l
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
U.S. EPA
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
1.6 ug/l
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
2-4 ug/l
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
10-20 ug/l
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
20-2000 ug/l
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.19\columnwidth}\raggedright\strut
\textgreater{}2000 ug/l
\strut\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.11\columnwidth}\raggedright\strut
Source
\strut\end{minipage} &
\begin{minipage}[b]{0.16\columnwidth}\raggedright\strut
Type
\strut\end{minipage} &
\begin{minipage}[b]{0.17\columnwidth}\raggedright\strut
Microcystin
\strut\end{minipage} &
\begin{minipage}[b]{0.17\columnwidth}\raggedright\strut
Chlorophyll
\strut\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
U.S. EPA
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
0.3
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
20.8
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
1
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
71.2
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
U.S. EPA
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
1.6
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
92.16
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
2
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
110.2
\strut\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

>>>>

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use and managing water bodies to meet
those levels could have far reaching benefits. However, compared to
other water quality measures, measurements of microcystin are not common
and current field measurement techniques have limited precision and
accuracy. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper we use this association to provide estimates
of chlorophyll \emph{a} that if exceeded would be indicative of a higher
probability of exceeding select health advisory concentrations for
microcystin-LR. Using the 2007 National Lakes Assessment and a
conditional probability approach that has been used in other water
quality settings, we identify chlorophyll \emph{a} concentrations that
are more likely than not to be associated with an exceedance of a
microcystin health advisory level. We look at the recent US EPA
standards for drinking water as well as the World Health Organization
levels for drinking water and recreational use. For microcystin
concentrations of 0.3, 1, 1.6, and 2 we find chlorophyll \emph{a}
concentrations of 20.8, 78.29, 109.66, and 109.66, respectively. When
managing for these various microcystin levels exceeding these reported
chlorophyll \emph{a} concentrations should be a trigger for further
testing and possible management action.

@@

\subsection{Data}\label{data}

\subsection{Conditional Probability
Analysis}\label{conditional-probability-analysis}

We used a conditional probability analysis approach to explore
associations between chlorophyll \emph{a} concentrations and World
Health Organization (WHO) and U.S. Environmental Protection Agency (U.S.
EPA) microcystin-LR health advisory levels {[}REFS{]}. Many levels have
been suggested {[}Table \ref{tab:microcystin_levels}{]}, but lakes with
higher microcystin-LR concentrations in the NLA were rare. Only 1.16 \%
of lakes sampled had a concentration greater than 10. Thus, for this
analysis we focus on the microcystin concentrations that are better
represented in the NLA data. These were 0.3, 1, 1.6, and 2 \(\mu\)g/L.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/chla_micro_scatter-1.pdf}
\caption{Scatterplot showing association betweeen chlorophyll
\it{a} and microcystin-LR. \label{fig:chla_micro_scatter}}
\end{figure}

@@

\begin{verbatim}
## No id variables; using all as measure variables
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/chla_micro_cdf-1.pdf}
\caption{Conditional probability plots showing association between the
probability of exceeding various microcystin-LR (MLR) health advisory
Levels. \label{fig:chla_micro_cdf}}
\end{figure}

@@

\begin{verbatim}
## Loading required package: grid
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/epa_child_cp_plot-1.pdf}
\caption{Conditional probability plots showing association between the
probability of exceeding various microcystin-LR (MLR) health advisory
Levels. \label{fig:multi_cp_plot}}
\end{figure}

@@

\begin{longtable}[c]{@{}lll@{}}
\toprule
\begin{minipage}[b]{0.11\columnwidth}\raggedright\strut
Source
\strut\end{minipage} &
\begin{minipage}[b]{0.16\columnwidth}\raggedright\strut
Type
\strut\end{minipage} &
\begin{minipage}[b]{0.20\columnwidth}\raggedright\strut
Concentration
\strut\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
1 \(\mu\)g/L
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
U.S. EPA
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
0.3 \(\mu\)g/L
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
U.S. EPA
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
1.6 \(\mu\)g/L
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
2-4 \(\mu\)g/L
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
10-20 \(\mu\)g/L
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
20-2000 \(\mu\)g/L
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.20\columnwidth}\raggedright\strut
\textgreater{}2000 \(\mu\)g/L
\strut\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.11\columnwidth}\raggedright\strut
Source
\strut\end{minipage} &
\begin{minipage}[b]{0.16\columnwidth}\raggedright\strut
Type
\strut\end{minipage} &
\begin{minipage}[b]{0.17\columnwidth}\raggedright\strut
Microcystin
\strut\end{minipage} &
\begin{minipage}[b]{0.17\columnwidth}\raggedright\strut
Chlorophyll
\strut\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
U.S. EPA
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
0.3
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
20.8
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
1
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
78.29
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
U.S. EPA
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Drinking
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
1.6
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
109.7
\strut\end{minipage}\tabularnewline
\begin{minipage}[t]{0.11\columnwidth}\raggedright\strut
WHO
\strut\end{minipage} &
\begin{minipage}[t]{0.16\columnwidth}\raggedright\strut
Recreational
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
2
\strut\end{minipage} &
\begin{minipage}[t]{0.17\columnwidth}\raggedright\strut
109.7
\strut\end{minipage}\tabularnewline
\bottomrule
\end{longtable}
